A Mechanised Library for Generic (Co)Programming

### 2024-02-26 paper/paper.tex
add comment to section 1 about establishing the main goals and contributions of the paper
Stuff

The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.
\end{enumerate}
In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

@@

A container is defined by a type of \emph{shapes}, and a \emph{family of
position types} indexed by shapes.
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Shape -> Set).
\end{minted}
An \emph{extension} of this container is a functor, whose action on objects is
given by:
\begin{minted}{Coq}
Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape.
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
Record App (X : Set) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

Definition Ks (A : Set) := A.
Definition Kp := False.
Instance Const A : Cont (Ks A) Kp := {
  valid := fun _ p =>
    match p with end }.

@@

These definitions are isomorphic to the equivalent container extension.

We illustrate how these definitions
\paragraph*{Isomorphisms with Container Extensions} Our definitions involving
polynomial types are isomorphic to polynomial functors as container extensions.

@@

>>>>

\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper. Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.
\end{enumerate}
In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

@@

A container is defined by a type of \emph{shapes}, and a \emph{family of
position types} indexed by shapes.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Shape -> Type).
\end{minted}
An \emph{extension} of this container is a functor, whose action on objects is
given by:
\begin{minted}{Coq}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

Definition Ks (A : Type) := A.
Definition Kp := False.
Instance Const A : Cont (Ks A) Kp := {
  valid := fun _ p =>
    match p with end }.

@@

\subsubsection{Isomorphisms of Coq Types and Container Extensions}

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : equiv X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Note the use of \mintinline{Coq}{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \mintinline{Coq}{App F X} is isomorphic to \mintinline{Coq}{A}, and
\mintinline{Coq}{App G X} is isomorphic to \mintinline{Coq}{B}. Then we can
show that \mintinline{Coq}{App (Prod F G) X} is isomorphic to
\mintinline{Coq}{A * B}.  If we have an element of type 
\mintinline{Coq}{App (Prod F G) X},
using the \mintinline{Coq}{inl} position, we can obtain
\mintinline{Coq}{App F X}.  Similarly, using \mintinline{Coq}{inr}, we can
obtain \mintinline{Coq}{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \mintinline{Coq}{App F X} and \mintinline{Coq}{App G X} to
obtain \mintinline{Coq}{A * B}. Similarly if we have \mintinline{Coq}{A * B},
we can first use the isomorphisms of 
\mintinline{Coq}{A} and
\mintinline{Coq}{B} to obtain
\mintinline{Coq}{App F X * App G X}, and then construct the necessary container
extension. Given \mintinline{Coq}{p_inl}/\mintinline{Coq}{p_inr} that act as
\mintinline{Coq}{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{minted}{Coq}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{minted}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that Product containers only have two valid positions.

@@

Since our framework only allows types with equivalences, and proper morphisms,
we can leverage Coq's \emph{setoid rewrite}

\dcas{Blabla: examples, no need to explain setoid rewrite in detail here}
